Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-24 paper-new/appendix.tex
work on proofs in appendix

\begin{lemma}\label{lem:step-1-model-to-step-2-model}
Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model} with dyn.

Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model} with dyn.
\end{lemma}
\begin{proof}
    % Write 
    % %
    % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
    % %

    Define a step-2 model $\mathcal M'$ as follows:
    \begin{itemize}
      \item Value objects are tuples consisting of:
      \begin{itemize}
        \item A value object $A$ in $\vf$ 
        \item A monoid of ``pure'' perturbations $P_A$ 
        \item A homomorphism of monoids $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$
        \item A monoid of ``impure'' perturbations $P^K_A$ that contains a distinguished element $\delta^*$
        \item A homomorphism of monoids $\ptbk_A : P^K_A \to \{ \phi \in \ef(FA, FA) \mid \phi \bisim \id_{FA} \}$
        such that $\ptbk_A(\delta^*) = \delta_A^*$
      \end{itemize}

      \item Computation objects are tuples consisting of:
      \begin{itemize}
        \item A computation object $B$ in $\ef$
        \item A monoid of ``pure'' perturbations $P_B$
        \item A homomorphism of monoids $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$
        \item A monoid of ``impure'' perturbations $P^K_B$
        \item A homomorphism of monoids $\ptbk_B : P^K_B \to \{ g \in \vf(UB, UB) \mid g \bisim \id_{UB} \}$.
      \end{itemize}

      \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively
      %, i.e.,
      % \[ \vf'((A, P_A, \ptb_A, P^K_A, \ptbk_A), (A', P_{A'}, \ptb_{A'}, P^K_{A'}, \ptbk_{A'})) = \vf(A, A') \]
      %
      % and likewise for computations.

@@

    Before introducing the relations, we make a definition.

    \begin{definition}[push-pull structure]
      Let $c : A \rel A'$ be a value relation of $\mathcal M$. A \emph{value push-pull structure} $\piv_c$ for $c$ consists of:
      \begin{itemize}
        \item A function $\push : P_A \to P_{A'}$ 
              such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
        \item A function $\push^K : P^K_A \to P^K_{A'}$ 
              such that for all $\delta^K_l \in P^K_A$ we have $\delta^K_l \ltdyn_{Fc}^{Fc} \push(\delta^K_l)$.
        \item A function $\pull : P_{A'} \to P_A$
              such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
        \item A function $\pull^K : P^K_{A'} \to P^K_A$
              such that for all $\delta^K_r \in P^K_{A'}$ we have $\pull(\delta^K_r) \ltdyn_{Fc}^{Fc} \delta^K_r$.
      \end{itemize}

      For $d : B \rel B'$ a computation relation, we define a \emph{computation push-pull structure} $\pie_d$ for $d$
      in an analogous manner.
    \end{definition}

    Now we continue with the description of the construction:
    \begin{itemize}

      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
      \begin{itemize}
        \item A value relation $c \in \vsq$
        \item A push-pull structure $\piv_c$ for $c$
      \end{itemize}

      The objects of $\esq'$ are defined analogously.

@@

      % Functors \times, +, F, U, arrow

@@

      % \item We define $(A, \pv_A, \ptbv_A) \arr (B, \pe_B, \ptbe_B) = (A \arr B, \pv_A \times \pe_B, h_\arr)$
      % where $\times$ is the product in the category of monoids, and $h_\arr$ is defined by 
      % $h_\arr(p_A, p_B) = \ptbv_A(p_A) \arr \ptbe_B(p_B)$.
    \end{itemize}
\end{proof}

@@

@@

  Then we can construct a \hyperref[def:step-3-model]{step-3 intensional model}.
\end{lemma}
\begin{proof}
  Write 
  %
  \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
  %

  We begin with a definition.

@@

  Likewise, let $d : B \rel B'$. A \emph{right-representation structure} $\rho^R_d$ for $d$ consists of
  a computation morphism $p_d \in \ef(B', B)$ such that $d$ is quasi-right-representable by $p_d$ (see Definition \ref{def:quasi-right-representable}).
  \end{definition}

@@

  We define a step-3 model $\mathcal M'$ as follows:
  \begin{itemize}
    \item The objects of $\mathcal M'$ are defined to be the same as the objects of $\mathcal M$.
    \item The value and computation morphisms in $\mathcal M'$ are the same as those of $\mathcal M$.
    \item A value relation is defined to be a tuple $(c, \rho^L_c, \rho^R_{Fc})$ where:
    \begin{itemize}
      \item $c$ is a value relation in $\mathcal M$, and 
      \item $\rho^L_c$ a left-representation structure for $c$, and 
      \item $\rho^R_{Fc}$ a right-representation structure for $Fc$.
    \end{itemize}
    \item Likewise, a computation relation is defined to be a tuple $(d, \rho^R_d, \rho^L_{Ud})$ with
    \begin{itemize}
      \item $d$ a computation relation in $\mathcal M$
      \item $\rho^R_d$ a right-representation structure for $d$
      \item $\rho^L_{Ud}$ a left-representation structure for $Ud$
    \end{itemize}
    \item Morphisms of value relations (i.e., the value squares) are defined by simply
    ignoring the representation structures. That is, a morphism of value relations
    $\alpha \in \vsq'((c, \rho^L_c, \rho^R_{Fc}), (c' \rho^L_{c'}, \rho^R_{Fc'}))$ is simply a morphism of value
    relations in $\vsq(c, c')$. Likewise for computations.
  \end{itemize}
\end{proof}

% Now we define the functors $F$, $U$, $\times$, and $\arr$.

% On objects, the behavior is the same as the respective functors in $\mathcal M$.

% For relations, we define 
% $\Fsq' (c, \rho^L_c, \rho^R_{Fc}) = (\Fsq c, \rho^R_{Fc}, UF(\rho^L_c))$ and
% $\Usq' (d, \rho^R_d, \rho^L_{Ud}) = (\Usq d, \rho^L_{Ud}, FU(\rho^R_d))$.

>>>>

\subsection{Constructing a Model with Perturbations}

The goal of this section is to prove the following lemma:

\begin{lemma}\label{lem:step-1-model-to-step-2-model}
Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model}.

Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model}.
\end{lemma}

We begin with a definition and some lemmas that will be useful in the construction
of the model.

\begin{definition}
  Let $c : A \rel A'$ be a value relation of $\mathcal M$.
  Let $P_A$ be a monoid of perturbations on $A$ and $P_{A'}$ a monoid of perturbations on $A'$.
  A \emph{push-pull structure} $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$ consists of:
  \begin{itemize}
    \item A function $\push : P_A \to P_{A'}$ 
    such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
    \item A function $\pull : P_{A'} \to P_A$
    such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
  \end{itemize}

  We define a push-pull structure $\pie_d$ for $d : B \rel B'$ with respect to perturbation monoids
  $P_B$ and $P_{B'}$ in an analogous manner.

  % \eric{do we need that $\delta^*$ is preserved by push and pull?}
  % We stipulate the additional requirement that if $\delta_A^*$ is in $P_B$ 
  % (where $\delta_A^*$ is the distinguished perturbation required to be in every $\ef(FA, FA)$),
  % then $\push(\delta_A^*) = \delta_{A'}^*$ and likewise $\pull(\delta_{A'}^*) = \delta_A^*$.
\end{definition}

%%%%%%%%%%%%%%%
% Composition %
%%%%%%%%%%%%%%%
\begin{lemma}
  Let $c : A \rel A'$ and $c' : A' \rel A''$ be value relations,
  and let $P_A, P_{A'}$ and $P_{A''}$ be monoids of perturbations.
  Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$, and
  and a push-pull structure $\piv_{c'}$ for $c'$ with respect to $P_{A'}$ and $P_{A''}$,
  we can define a push-pull structure $\piv_{c \comp c'}$ for $c \comp c'$.

  Likewise, we can define a push-pull structure for the composition of computation relations.
\end{lemma}
\begin{proof}
  We define $\piv_{c \comp c'}$ as the following push-pull structure:

  \begin{itemize}
    \item $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$
    \item $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$
  \end{itemize}

  We observe that the required squares exist for both push and pull.
  In particular, for push we have that $\delta^l \ltdyn_c^c \push_c(\delta^l)$
  using the push property for $c$, and then using the push property for $c'$ we have
  $\push_c(\delta^l) \ltdyn_{c'}^{c'} \push_{c'}(\push_c(\delta^l))$.
  We can then compose these squares horizontally to obtain the desired square.
  The pull property follows similarly.

  The push-pull structure for the composition of computation relations is defined analogously.
\end{proof}

%%%%%%%%%%%%%
% UF and FU %
%%%%%%%%%%%%%
\begin{lemma}
  Let $c : A \rel A'$ and let $P_A$ and $P_{A'}$ be monoids of perturbations.
  Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$,
  we can define a push-pull structure $\pie_{UF(c)}$ for $UF(c)$ with respect to $UF(P_{A})$
  and $UF(P_{A'})$.
  % Eric: Are we assuming here that F and U preserve perturbations?
\end{lemma}
\begin{proof}
  Follows by the functorial action of $UF$ and $FU$ on perturations and squares.
\end{proof}

%%%%%%%%%%%%
% Products %
%%%%%%%%%%%%
\begin{lemma}
  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, and let $P_{A_1}$,
  $P_{A_2}$, $P_{A_1'}$ and $P_{A_2'}$ be monoids of perturbations.
  Given push-pull structures 
  $\piv_{c_1}$ with respect to $P_{A_1}$ and $P_{A_1'}$, and 
  $\piv_{c_2}$ with respect to $P_{A_2}$ and $P_{A_2'}$,
  we can define a push-pull structure $\piv_{c_1 \times c_2}$
  with respect to $P_{A_1} \times P_{A_2}$ and $P_{A_1'} \times P_{A_2'}$.

  Likewise, given push-pull structures 
  $\pie_{Fc_1}$ with respect to $F(P_{A_1})$ and $F(P_{A_1'})$, and
  $\pie_{Fc_2}$ with respect to $F(P_{A_2})$ and $F(P_{A_2'})$,
  we can define a push-pull structure $\pie_{F(c_1 \times c_2)}$
  with respect to $F(P_{A_1} \times P_{A_2})$ and $F(P_{A_1'} \times P_{A_2'})$.
\end{lemma}
\begin{proof}
We define the push function for $\piv_{c_1 \times c_2}$ by 
$\push(\delta_1, \delta_2) = (\push_{c_1}(\delta_1), \push_{c_2}(\delta_2))$ and likewise for $\pull$.
The push-pull property holds because it holds for each component.
\end{proof}

%%%%%%%%%
% Arrow %
%%%%%%%%%
\begin{lemma}
  Let $c : A \rel A'$ and $d : B \rel B'$.
  Given push-pull structures $\piv_c$ for $c$ and $\pie_d$ for $d$,
  we can define a push-pull structure $\pie_{c \arr d}$ for $c \arr d$.

  Given push-pull structures $\pie_{Fc}$ for $Fc$ and $\piv_{Ud}$ for $Ud$,
  we can define a push-pull structure $\piv_{U(c \arr d)}$ for $U(c \arr d)$
  with respect to the monoids $(P_{FA})^{op} \oplus P_B$
\end{lemma}
\begin{proof}
  The former follows from the functorial action of $\arr$ on perturbations and squares,
  and the latter uses the functorial action of $\tok$ on perturbations and squares.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%
% Model Construction %
%%%%%%%%%%%%%%%%%%%%%%
We now proceed with the construction of the model:
    % Write 
    % %
    % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
    % %

    Define a step-2 model $\mathcal M'$ as follows:
    \begin{itemize}
      \item Value objects are triples consisting of:
      \begin{itemize}
        \item A value object $A$ in $\vf$ 
        \item A monoid $P_A$ of ``pure'' perturbations that is a submonoid of 
        $\{ f \in \vf(A, A) \mid f \bisim \id_A \}$
        \item A monoid $P^K_A$ of ``impure'' or ``Kleisli'' perturbations that is a submonoid of
        $\{ \phi \in \ef(FA, FA) \mid \phi \bisim \id_{FA} \}$ such that $P^K_A$ contains
        $\delta_A^*$
      \end{itemize}

      \item Computation objects are triples consisting of:
      \begin{itemize}
        \item A computation object $B$ in $\ef$
        \item A monoid $P_B$ of ``pure'' perturbations that is a submonoid of
        $\{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$
        \item A monoid $P^K_B$ of ``impure'' or ``Kleisli'' perturbations that is a submonoid
        of $\{ g \in \vf(UB, UB) \mid g \bisim \id_{UB} \}$
      \end{itemize}

      \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively.
      %, i.e.,
      % \[ \vf'((A, P_A, \ptb_A, P^K_A, \ptbk_A), (A', P_{A'}, \ptb_{A'}, P^K_{A'}, \ptbk_{A'})) = \vf(A, A') \]
      %
      % and likewise for computations.

@@

    % Before introducing the relations, we make a definition.

    % \begin{definition}[push-pull structure]
    %   Let $c : A \rel A'$ be a value relation of $\mathcal M$. A \emph{value push-pull structure} $\piv_c$ for $c$ consists of:
    %   \begin{itemize}
    %     \item A function $\push : P_A \to P_{A'}$ 
    %           such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
    %     \item A function $\push^K : P^K_A \to P^K_{A'}$ 
    %           such that for all $\delta^K_l \in P^K_A$ we have $\delta^K_l \ltdyn_{Fc}^{Fc} \push(\delta^K_l)$.
    %     \item A function $\pull : P_{A'} \to P_A$
    %           such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
    %     \item A function $\pull^K : P^K_{A'} \to P^K_A$
    %           such that for all $\delta^K_r \in P^K_{A'}$ we have $\pull(\delta^K_r) \ltdyn_{Fc}^{Fc} \delta^K_r$.
    %   \end{itemize}

    %   For $d : B \rel B'$ a computation relation, we define a \emph{computation push-pull structure} $\pie_d$ for $d$
    %   in an analogous manner.
    % \end{definition}

    % Now we continue with the description of the construction:
    \begin{itemize}

      \item The objects of $\vsq'$ (i.e., the value relations) are triples consisting of:
      \begin{itemize}
        \item A value relation $c \in \vsq$.
        \item A push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$.
        \item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
      \end{itemize}

      The objects of $\esq'$ are defined analogously, i.e.,
      \begin{itemize}
        \item A computation relation $d \in \esq$.
        \item A push-pull structure $\pie_d$ for $d$ with respect to $P_B$ and $P_{B'}$.
        \item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
      \end{itemize}

@@

      \item We define composition of relations $(c, \piv_c, \pie_{Fc})$ and 
      $(c', \piv_{c'}, \pie_{Fc'})$ as $(c \comp c', \piv_{c \comp c'}, \pie_{Fc \comp Fc'})$
      and likewise for computation relations.

      % Functors \times, +, F, U, arrow
      Now we define the actions of the functors:
      \begin{itemize}

      % \item We begin with the functor $m$ that composes relations and squares horizontally.
      % Given value relations $(c ,\piv_c, \pie_{Fc})$ and $(c', \piv_{c'}, \pie_{Fc'})$ where $c : A \rel A'$ and $c' : A' \rel A''$,
      % we define their composition to be the triple $(c \comp c', \piv_{c \comp c'})$ where we define
      % $\piv_{c \comp c'}$ as the following push-pull structure:

      % \begin{itemize}
      %   \item $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$
      %   \item $\push^K_{c \comp c'} = \push^K_{c'} \circ \push^K_{c}$
      %   \item $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$
      %   \item $\pull^K_{c \comp c'} = \pull^K_{c} \circ \pull^K_{c'}$
      % \end{itemize}

      % We observe that each of the components satisfies the commuting necessary square.

      \item We define $\times$ on objects by

      \[ (A_1, P_{A_1}, P^K_{A_1}) \times (A_2, P_{A_2}, P^K_{A_2}) =
        (A_1 \times A_2, P_{A_1} \times P_{A_2}, P^K_{A_1} \oplus P^K{A_2}). \]

      where $\oplus$ is the coproduct in the category of monoids.

      We define $\times$ on relations by

      \[ (c_1, \piv_{c_1}, \pie_{Fc_1}), (c_2 \piv_{c_2}, \pie_{Fc_2}) = 
        (c_1 \times c_2, \piv_{c_1 \times c_2}, \pie_{F(c_1 \times c_2)}). \]

      \item We define $F$ on objects by

      \[ F(A, P_A, P^K_A) = (FA, P^K_A, UF(P_A)). \]

      We define $F$ on relations by

      \[ F(c, \piv_c, \pie_{Fc}) = (Fc, \pie_{Fc}, \piv_{UF(c)}). \]

      \item We define $U$ on objects by

      \[ U(B, P_B, P^K_B) = (UB, P^K_B, FU(P_B). )\]

      We define $U$ on relations by

      \[ U(d, \pie_d, \piv_{Ud}) = (Ud, \piv_{Ud}, \pie_{FU(d)})  \]

      We define $\arr$ on objects by

      \[ (A, P_A, P^K_A) \arr (B, P_B, P^K_B) = 
        (A \arr B, (P_A)^{op} \times P_B, (P^K_A)^{op} \oplus P^K_B). \]

      We define $\arr$ on relations by

      \[ (c, \piv_c, \pie_{Fc}) \arr (d, \pie_d, \piv_{Ud}) =
        (c \arr d, \pie_{c \arr d}, \piv_{U(c \arr d)}). \]

@@

      % \item We define $(A, \pv_A, \ptbv_A) \arr (B, \pe_B, \ptbe_B) = (A \arr B, \pv_A \times \pe_B, h_\arr)$
      % where $\times$ is the product in the category of monoids, and $h_\arr$ is defined by 
      % $h_\arr(p_A, p_B) = \ptbv_A(p_A) \arr \ptbe_B(p_B)$.
    \end{itemize}

@@

\subsection{Constructing a Model with Quasi-Representable Relations}

The goal of this section is to prove the following lemma:

@@

  Then we can construct a \hyperref[def:step-3-model]{step-3 intensional model}.
\end{lemma}

Before proceeding with the proof, we begin with a definition.

@@

Likewise, let $d : B \rel B'$ be a computation relation. A \emph{right-representation structure}
$\rho^R_d$ for $d$ consists of a computation morphism $p_d \in \ef(B', B)$
such that $d$ is quasi-right-representable by $p_d$ (see Definition
\ref{def:quasi-right-representable}).
\end{definition}

@@

% TODO establish the notation

%%%%%%%%%%%%%%%
% Composition %
%%%%%%%%%%%%%%%
\begin{lemma}\label{lem:representation-comp}
  In the below, let $c : A \rel A'$ and $c' : A' \rel A''$ and $d : B \rel B'$ and $d' : B' \rel B''$.
  \begin{enumerate}
    \item Given left-representation structures $\rho^L_c$ for $c$ and $\rho^L_{c'}$ for $c'$,
    we can define a left-representation structure for the composition $c \comp c'$
    \item Given right-representation structures $\rho^R_d$ for $d$ and $\rho^R_{d'}$ for $d'$,
    we can define a right-representation structure for the composition $d \comp d'$
    \item Given right-representation structures $\rho^R_{Fc}$ for $Fc$ and $\rho^R_{Fc'}$ for $Fc'$,
    we can define a right-representation structure $\rho^R_{F(c \comp c')}$ for $F(c \comp c')$.
    \item Given left-representation structures $\rho^L_{Ud}$ for $Ud$ and $\rho^L_{Ud'}$ for $Ud'$,
    we can define a left-representation structure $\rho^L_{U(d \comp d')}$ for $U(d \comp d')$.
  \end{enumerate}
  \eric{Do we need parts 3 and 4?}

\end{lemma}
\begin{proof}
  \begin{enumerate}

    % 1.
    \item We define $\rho^L_{c \comp c'}$ as follows. In the definitions of the perturbations,
    we make use of the fact that $c$ and $c'$ satisfy the push-pull property.
    \begin{itemize}
      \item $e_{c \comp c'} = e_{c'} \circ e_c$
      \item $\delre_{c \comp c'} = \delre_{c'} \circ \push_{c'}(\delre_c)$
      \item $\delle_{c \comp c'} = \pull_c(\delle_{c'}) \circ \delle_c$
      \item $\upl$ is the following square: \input{squares/UpL-comp}

      The square $(*)$ exists by the push-pull property for $c'$, and the square $(**)$ exists
      because $r$ is a unit for horizontal composition, so $r(A') \comp c' = c'$, and so this
      is simply the identity square $\id_{c'} \in \vsq(c', c')$.

      \item $\upr$ is the following square: \input{squares/UpR-comp}
    \end{itemize}

    % 2.
    \item We define $\rho^R_{d \comp d'}$ as follows:
    \begin{itemize}
      \item $p_{d \comp d'} = p_d \circ p_{d'}$
      \item $\dellp_{d \comp d'} = \dellp_d \circ \pull_d(\dellp_{d'})$
      \item $\delrp_{d \comp d'} = \push_{d'}(\delrp_d) \circ \delrp_{d'}$
      \item $\dnr$ is the following square: \input{squares/DnR-comp}

      Here the square $(*)$ exists by the push-pull property for $d$.

      \item $\dnl$ is the following square: \input{squares/DnL-comp}
    \end{itemize}

    % 3.
    \item We define $\rho^R_{F(c \comp c')}$ as follows:
    \begin{itemize}
      \item $p_{F(c \comp c')} = p_{Fc} \circ p_{Fc'}$ % : FA'' \to FA
      \item $\dellp_{F(c \comp c')} = $
      \item $\delrp_{F(c \comp c')} = $
    \end{itemize}

    % 4.
    \item We define $\rho^L{(U(d \comp d'))}$ as follows:

  \end{enumerate}
\end{proof}

% Want to show: U(d \comp d') is weakly equivalent to U(d) \comp U(d').
% This holds because both are quasi-representable by the same projection
% \begin{lemma}
%   Let $\mathcal M$ be a step-3 intensional model, and let
%   $d : B \rel B'$ and $d' : B' \rel B''$.
%   Let $\rho_\text{comp}$ be an arbitrary right-representation for $d \comp d'$.
%   Then the projection $p_{d \circ d'}$ is weakly equivalent to
% \end{lemma}

%%%%%%%%%%%%%
% UF and FU %
%%%%%%%%%%%%%
\begin{lemma}\label{lem:representation-UF-FU}
  Let $c : A \rel A'$, and let $\rho^L_c$ be a left-representation structure for $c$.
  Then we can define a left-representation structure $\rho^L_{UF(c)}$ for $UF(c)$.

  Similarly, let $d : B \rel B'$ and let $\rho^R_d$ be a right-representation structure for $d$.
  Then we can define a right-representation structure $\rho^R_{FU(d)}$ for $FU(d)$.
\end{lemma}
\begin{proof}
  We define $\rho^L_{UF(c)}$ as follows:
  \begin{itemize}
    \item $e_{UF(c)} = UF(e_c)$
    \item $\delre_{UF(c)} = UF(\delre_c)$ (which is in the monoid of perturbations of $UF(A')$ because the
    perturbation monoids are closed under the actions of the functors $F$ and $U$)
    \item $\delle_{UF(c)} = UF(\delle_c)$
    \item We get the two commuting squares by the functorial action of $UF$ on the two squares for $c$,
    i.e., $\upr_{UF(c)} = UF(\upr_c)$
  \end{itemize}

  We define $\rho^R_{FU(d)}$ in a similar manner.
\end{proof}

%%%%%%%%%%%%
% Products %
%%%%%%%%%%%%
\begin{lemma}\label{lem:representation-product}
  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$.
  Let $\rho^L_{c_1}$ be a left-representation structure for $c_1$, and
  let $\rho^L_{c_2}$ be a left-representation structure for $c_2$.
  Then we can define a left-representation structure for $c_1 \times c_2$.

  Likewise, let $\rho^R_{Fc_1}$ and $\rho^R_{Fc_2}$ be right-representation
  structures for $Fc_1$ and $Fc_2$ respectively.
  Then we can deinfe a right-representation structure for $F(c_1 \times c_2)$.
\end{lemma}
\begin{proof}
  We define $\rho^L_{c_1 \times c_2}$ as follows:
  \begin{itemize}
    \item $e_{c_1 \times c_2} = e_{c_1} \times e_{c_2}$
    \item $\delre_{c_1 \times c_2} = \delre_{c_1} \times \delre_{c_2}$ and
    likewise for $\delle_{c_1 \times c_2}$
    \item We get the commuting squares via the functorial action of $\times$
    on the corresponding squares for $c_1$ and $c_2$.
  \end{itemize}

  % TODO check this
  We define $\rho^R_{F(c_1 \times c_2)}$ as follows:
  \begin{itemize}
    \item $p_{F(c_1 \times c_2)} = (p_{Fc_1} \timesk A_2) \circ (A_1' \timesk p_{Fc_2})$
    \item $\dellp_{F(c_1 \times c_2)} = (\dellp_{Fc_1} \timesk A_2) \circ (A_1 \timesk \dellp_{Fc_2})$
    \item $\delrp_{F(c_1 \times c_2)} = (\delrp_{Fc_1} \timesk A_2') \circ (A_1' \timesk \delrp_{Fc_2})$
    \item The commuting squares are obtained via the functorial action of $\timesk$ on the squares for $Fc_1$ and $Fc_2$.
    % TODO elaborate?
  \end{itemize}
\end{proof}

%%%%%%%%%
% Arrow %
%%%%%%%%%
\begin{lemma}\label{lem:representation-arrow}
  Let $c : A \rel A'$ and $d : B \rel B'$.
  Let $\rho^L_c$ be a left-representation structure for $c$,
  and let $\rho^R_d$ be a right-representation structure for $d$.
  Then we can define a right-representation structure for $c \arr d$.

  Likewise, let $\rho^R_{Fc}$ be a right-representation structure for $Fc$,
  and let $\rho^L_{Ud}$ be a left-representation structure for $Ud$.
  Then we can define a left-representation structure for $U(c \arr d)$.
\end{lemma}
\begin{proof}
  We define $\rho^R_{c \arr d}$ as follows:
  \begin{itemize}
    \item $p_{c \arr d} = e_c \arr p_d \in \ef(A' \arr B', A \arr B)$ 
    (using the functorial action of $\arr$ on morphisms).
    \item $\dellp_{c \arr d} = \delle_c \arr \dellp_d$
    \item $\delrp_{c \arr d} = \delre_c \arr \delrp_d$
    \item The squares $\dnr$ and $\dnl$ are obtained via the functorial action of $\arr$, i.e., we define

    \[ \dnr_{c \arr d} = \upr_{c} \arr \dnr_{d} : 
      (\delle_c \arr \dellp_d) \ltdyn_{r(A \arr B)}^{c \arr d} (e_c \arr p_d),  \]

    and

    \[ \dnl_{c \arr d} = \upl_{c} \arr \dnl_{d}. \]

  % TODO check this
  We define $\rho^L_{U(c \arr d)}$ as follows:
  \begin{itemize}
    \item $e_{U(c \arr d)} = (p_{Fc} \tok B') \circ (A \tok e_{Ud})$
    \item $\delre_{U(c \arr d)} = (\delrp_{Fc} \tok B') \circ (A' \tok \delre_{Ud})$
    \item $\delle_{U(c \arr d)} = (\dellp_{Fc} \tok B) \circ (A \tok \delle_{Ud})$
    \item The squares $\upl$ and $\upr$ are obtained via the functorial action of $\tok$.
    For instance, $\upl$ is given by the following square:

    % https://q.uiver.app/#q=WzAsNixbMCwwLCJVKEEgXFx0byBCKSJdLFsyLDAsIlUoQScgXFx0byBCJykiXSxbMiwxLCJVKEEnIFxcdG8gQicpIl0sWzIsMiwiVShBJyBcXHRvIEInKSJdLFswLDEsIlUoQSBcXHRvIEInKSJdLFswLDIsIlUoQScgXFx0byBCJykiXSxbMCwxLCJVKGMgXFx0byBkKSJdLFs0LDIsIlUoYyBcXHRvIHIoQicpKSJdLFs1LDMsIlUocihBJykgXFx0byByKEInKSkiXSxbNCw1LCJwX3tGY30gXFx0b2sgQiciLDJdLFsxLDIsIkEnIFxcdG9rIFxcZGVscmVfe1VkfSJdLFsyLDMsIlxcZGVscnBfe0ZjfSBcXHRvayBCJyJdLFswLDQsIkEgXFx0b2sgZV97VWR9IiwyXSxbMTIsMTAsIlxcaWRfe0ZjfSBcXHRvayBcXHVwbF97VWR9IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzksMTEsIlxcZG5sX3tGY30gXFx0b2sgXFxpZF97cihCJyl9IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
    \[\begin{tikzcd}[ampersand replacement=\&,row sep=large]
      {U(A \to B)} \&\& {U(A' \to B')} \\
      {U(A \to B')} \&\& {U(A' \to B')} \\
      {U(A' \to B')} \&\& {U(A' \to B')}
      \arrow["{U(c \to d)}", from=1-1, to=1-3]
      \arrow["{U(c \to r(B'))}", from=2-1, to=2-3]
      \arrow["{U(r(A') \to r(B'))}", from=3-1, to=3-3]
      \arrow[""{name=0, anchor=center, inner sep=0}, "{p_{Fc} \tok B'}"', from=2-1, to=3-1]
      \arrow[""{name=1, anchor=center, inner sep=0}, "{A' \tok \delre_{Ud}}", from=1-3, to=2-3]
      \arrow[""{name=2, anchor=center, inner sep=0}, "{\delrp_{Fc} \tok B'}", from=2-3, to=3-3]
      \arrow[""{name=3, anchor=center, inner sep=0}, "{A \tok e_{Ud}}"', from=1-1, to=2-1]
      \arrow["{\id_{Fc} \tok \upl_{Ud}}"{description}, draw=none, from=3, to=1]
      \arrow["{\dnl_{Fc} \tok \id_{r(B')}}"{description}, draw=none, from=0, to=2]
    \end{tikzcd}\]
  \end{itemize}

  The construction of $\upr$ is similar.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%
% Model Construction %
%%%%%%%%%%%%%%%%%%%%%%
Now we can give the proof of the main lemma:
  % Write 
  % %
  % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
  % %

  We define a step-3 model $\mathcal M'$ as follows:
  \begin{itemize}
    \item The objects of $\mathcal M'$ are defined to be the same as the objects of $\mathcal M$.
    \item The value and computation morphisms in $\mathcal M'$ are the same as those of $\mathcal M$.
    \item A value relation is defined to be a tuple $(c, \rho^L_c, \rho^R_{Fc})$ with
    \begin{itemize}
      \item $c$ a value relation in $\mathcal M$,
      \item $\rho^L_c$ a left-representation structure for $c$, and
      \item $\rho^R_{Fc}$ a right-representation structure for $Fc$
    \end{itemize}
    \item Likewise, a computation relation is defined to be a tuple $(d, \rho^R_d, \rho^L_{Ud})$ with
    \begin{itemize}
      \item $d$ a computation relation in $\mathcal M$,
      \item $\rho^R_d$ a right-representation structure for $d$, and
      \item $\rho^L_{Ud}$ a left-representation structure for $Ud$.
    \end{itemize}
    \item Morphisms of value relations (i.e., the value squares) are defined by simply
    ignoring the representation structures. That is, a morphism of value relations
    $\alpha \in \vsq'((c, \rho^L_c, \rho^R_{Fc}), (c' \rho^L_{c'}, \rho^R_{Fc'}))$ is simply a morphism of value
    relations in $\vsq(c, c')$. Likewise for computations.
  \end{itemize}

We define the functor $m$ which composes relations and squares horizontally as follows:
Let $c : A \rel A'$ and $c' : A' \rel A''$. We let

\begin{align*} 
  m ((c, \rho^L_c, \rho^R_{Fc}), (c', \rho^L_{c'}, \rho^R_{Fc'})) = \\
  (c \comp c', \rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}). 
\end{align*}

Now we define the functors $F$, $U$, $\times$, and $\arr$.
On objects, the behavior is the same as the respective functors in $\mathcal M$.
For relations, we define

\[ \Fsq^{\mathcal M'} (c, \rho^L_c, \rho^R_{Fc}) = 
  (\Fsq^M c, \rho^R_{Fc}, \rho^L_{UF(c)}), \]
and
\[ \Usq^{\mathcal M'} (d, \rho^R_d, \rho^L_{Ud}) = 
  (\Usq^M d, \rho^L_{Ud}, \rho^R_{FU(d)}), \]

where $\rho^L_{UF(c)}$ and $\rho^R_{FU(d)}$ are as defined in the proof of Lemma
\ref{lem:representation-UF-FU}.
